---
title: ""
output:
  html_document:
    theme: cosmo
    toc: false
    number_sections: false
    highlight: tango
    css: style.css
date: "`r format(Sys.Date(), '%B %d, %Y')`"
bibliography: references.bib
link-citations: true
citeproc: true
nocite: '@*'
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

library(data.table)
library(ggplot2)
library(scales)
library(gt)
library(rprojroot)

# Use project root (stable when knitting)
root_dir <- find_root(has_file("proposal.Rproj"))
knitr::opts_knit$set(root.dir = root_dir)

source(file.path(root_dir, "code/00_depends.R"))

# ------------------------------------------------------------
# Council bargaining unit data: Web
# ------------------------------------------------------------
data_dir <- file.path(root_dir, "data")
barg_path <- file.path(data_dir, "barg_units_group_levels.csv")
barg <- data.table::fread(barg_path)

barg[, level := data.table::fcase(
  tolower(trimws(level)) == "junior", "Base",
  tolower(trimws(level)) == "senior", "Senior",
  default = level
)]

web <- barg[group == "web"]

# Tenure in years (non-negative)
web[, tenure_years := as.numeric((fy_start - agency_start_date) / 365)]
web[tenure_years < 0, tenure_years := 0]

# ------------------------------------------------------------
# City comparable: Computer Specialist (Software) (Title 1362 / 13632 series)
# ------------------------------------------------------------
# NOTE: some NYC materials use 13632 for Computer Specialist (Software).
# If your dataset uses 1362, keep it; otherwise 13632 is the common DCAS code.
cs_codes <- c("1362", "13632")

csts <- data.table::fread(
  "https://data.cityofnewyork.us/resource/nzjr-3966.csv?$limit=99999"
)

cst_range <- csts[title %in% cs_codes]
```

## Web Developers

City Council Web Developers design, build, and maintain web applications that support legislative operations, transparency, and internal workflows. Typical work includes full-stack development, accessibility and security practices, deployment and maintenance, and collaboration with policy and data staff.

```{r}
webp <- web[
  , .(
    n = .N,
    median_base_salary = median(base_salary, na.rm = TRUE),
    mean_tenure_years  = mean(tenure_years, na.rm = TRUE)
  ),
  by = .(fiscal_year, level)
][order(fiscal_year, level)]

webp[, fiscal_year := as.integer(fiscal_year)]
```

```{r}
ggplot(webp, aes(x = fiscal_year, y = median_base_salary)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(breaks = sort(unique(webp$fiscal_year))) +
  scale_y_continuous(labels = scales::label_dollar()) +
  labs(
    title = "Median base salary for Web Developers",
    subtitle = "FY 2018–2025",
    x = "Fiscal year",
    y = "Median base salary ($)",
    caption = "Source: NYC Open Data, Citywide Payroll Data (Fiscal Year)."
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank()) +
  facet_wrap(~level)
```

```{r}
gt::gt(
  webp[fiscal_year == "2025"
    , .(
      `Fiscal Year` = fiscal_year,
      level,
      N = n,
      `Median Base Salary` = median_base_salary,
      `Average Tenure (years)` = round(mean_tenure_years, 2),
      by = level
    )
  ]
) %>% 
  gt::fmt_currency(columns = `Median Base Salary`, currency = "USD", decimals = 0)
```

------------------------------------------------------------------------



NYC civil service materials describe Computer Specialists (Software) as responsible for the analysis, design, development, implementation, enhancement, maintenance, and security of software systems, including applications and websites. [@dcas_css_exam_notice]


NYC Jobs postings for Computer Specialist (Software) II and Computer Specialist (Software), Level III list minimum qualifications requiring a baccalaureate degree (including or supplemented by 24 credits in computer science or a related field) and two years of satisfactory full-time software experience in designing, programming, debugging, maintaining, implementing, and enhancing software applications (plus leadership/complex project contribution language). [@nycjobs_css_level2; @nycjobs_css_level3]

Level I is excluded because it often functions as an entry/trainee tier, while Council web roles generally require independent delivery and ongoing application ownership.

Bargaining implication: City Computer Specialist (Software) ranges exceed Council web developer medians, indicating a persistent gap for comparable technical work.

```{r}


  cst_range_sub <- cst_range[asg_lvl %in% c("02", "03")]


  gt::gt(
    cst_range_sub[
      , .(
        Title = descr,
        `Assignment Level` = asg_lvl,
        `Min Salary` = min_rate,
        `Max Salary` = max_rate
      )
    ]
  ) |>
    gt::fmt_currency(columns = c(`Min Salary`, `Max Salary`), currency = "USD", decimals = 0)


```
